---
title: "Vectorized functions"
subtitle: "with custom tables and figures"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

tmtyro is optimized for a fast start, speeding users along a standard workflow made possible by tidytext and the tidyverse packages on which it's built. But it also offers a pathway for growth by providing vectorized functions for use in a larger ecosystem of tools and simplifying the transition to text mining for anyone familiar with those tools.

## Loading a corpus

Regardless of a user's starting point, tmtyro's main verbs for gathering and preparing a corpus will prove useful:

```{r}
#| message: false
library(dplyr)
library(ggplot2)
library(gt)
library(tmtyro)
corpus_dubliners <- get_gutenberg_corpus(2814) |> 
  load_texts(lemma = TRUE, pos = TRUE) |> 
  identify_by(part) |> 
  standardize_titles()

corpus_dubliners
```

## Adding columns with dplyr

Beyond this point, anyone familiar with common tidyverse packages like dplyr and ggplot2 might forge their own path. For instance, while tmtyro offers a selection of verbs for adding new columns---all helpfully beginning with `add_...()`---the dplyr way to add columns is with `mutate()`. To support this workflow, tmtyro offers functions for working with columns as vectors, too.

### Word count and frequency

The main path for adding a column of word frequencies is with the `add_frequency()` function. Users familiar with dplyr can instead use `mutate()` paired with `get_frequency()`:

```{r}
#| message: false
corpus_dubliners |> 
  select(doc_id, word, lemma) |> 
  mutate(
    count_word = get_frequency(word),
    count_lemma = get_frequency(lemma))
```

These frequencies can just as easily be reported as percentages relative to the whole, using `get_frequency(percent = TRUE)`:

```{r}
corpus_dubliners |> 
  select(doc_id, word, lemma) |> 
  mutate(
    count_word = get_frequency(word, percent = TRUE),
    count_lemma = get_frequency(lemma, percent = TRUE))
```

Of course, these frequencies relate to the entire corpus. To get document-level numbers, use the dplyr functions `group_by()` and `ungroup()` around `mutate()`:

```{r}
corpus_dubliners <- corpus_dubliners |> 
  select(doc_id, word) 

dubliners_count <- corpus_dubliners |> 
  group_by(doc_id) |> 
  mutate(
    n = get_frequency(word),
    freq = get_frequency(word, percent = TRUE)) |> 
  ungroup()

dubliners_count
```

### Vocabulary richness

tmtyro offers `add_vocabulary()` for adding columns devoted to vocabulary growth, uniqueness, and ratios of lexical diversity. When using `mutate()`, these features are handled by testing functions like `is_new()` and `is_hapax()` and measuring functions like `get_cumulative_vocabulary()`, `get_ttr()`, and `get_hir()`. As with word count, it's usually best to calculate values grouped by document using `group_by()` and `ungroup()`:

```{r}
dubliners_vocab <- corpus_dubliners |> 
  group_by(doc_id) |> 
  mutate(
    new_word = is_new(word), 
    hapax_word = is_hapax(word),
    vocab = get_cumulative_vocabulary(word), 
    ttr = get_ttr(word), 
    hir = get_hir(word)) |> 
  ungroup()

dubliners_vocab
```

Slower than these other methods, `get_htr()` offers more than is available in `add_vocabulary()`, returning the hapax-token ratio. This method can be slower, so be careful when applying it to a large corpus

```{r}
dubliners_vocab <- 
  dubliners_vocab |> 
  select(doc_id, word, ttr, hir) |> 
  filter(doc_id %in% c("The Sisters", "An Encounter")) |> 
  group_by(doc_id) |> 
  mutate(
    htr = get_htr(word)
  ) |> 
  ungroup()

dubliners_vocab |> 
  # skip the first few lines
  {\(x) x[-c(1:6),]}()
```

### Dictionary Matching

While `add_dictionary()` can manage adding columns matching a dictionary to a column like `word` or `lemma`, the `get_match()` function will return the dictionary match for any word or vector. It's used to add a column with dplyr's `mutate()` like this:

```{r}
emoji_weather <- make_dictionary(
  list(
    "ï¸â˜”ï¸" = c("rain", "rains", "rainy", "raining"),
    "ï¸â›ˆï¸" = c("storm", "storms", "stormy", "storming"),
    "â˜ï¸" = c("cloud", "clouds", "cloudy"),
    "ðŸŒž" = c("sun", "sunny"),
    "ðŸŒ«ï¸" = c("fog", "fogs", "foggy", "mist", "misty"),
    "ðŸŒ¬ï¸" = c("wind", "winds", "windy"),
    "ï¸â„ï¸" = c("snow", "snows", "snowing")),
  name = "weather")

dubliners_weather <- corpus_dubliners |> 
  mutate(weather = get_match(word, emoji_weather))

dubliners_weather |> 
  # show only one story and skip a few hundred words
  filter(doc_id == "The Dead") |> 
  filter(row_number() > 602)

dubliners_weather |> 
  drop_na()
```

#### Sentiment

Matching sentiment with `get_sentiment()` just uses a special kind of dictionary matching:

```{r}
corpus_dubliners |> 
  mutate(
    sent_1 = get_match(word, tidytext::get_sentiments("bing")),
    sent_2 = get_sentiment(word, "bing")) |> 
  drop_na()
```

### Tf-idf

The `get_tf()` function offers a shorthand alias to `get_frequency(percent = TRUE)`. Importantly, this frequency reports values for the whole of a group, so it may often be necessary to use `group_by()` and `ungroup()`:

```{r}
dubliners_tfidf <- corpus_dubliners |> 
  group_by(doc_id) |> 
  mutate(
    tf = get_tf(word)) |> 
  ungroup()

dubliners_tfidf
```

Simpler is to use `get_tf_by()`, which accepts a string and a grouping variable like `doc_id`. It works well with `get_idf_by()` to calculate the inverse document frequency for each term. Lastly, multiplying these columns together results in tf-idf:

```{r}
dubliners_tfidf <- corpus_dubliners |> 
  mutate(
    tf = get_tf_by(word, doc_id),
    idf = get_idf_by(word, doc_id),
    tf_idf = tf * idf)

dubliners_tfidf
```

Alternatively, `get_tfidf_by()` simplifies the process to one step:

```{r}
dubliners_tfidf |> 
  select(doc_id, word, tf_idf) |> 
  mutate(
    tf_idf2 = get_tfidf_by(word, doc_id)
  )
```

## Preparing tables with dplyr and gt

In addition to adding new columns, tmtyro's functions like `add_vocabulary()` and `add_sentiment()` also prepare objects to work easily with `tabulize()`. Users who wish to prepare similar tables manually will need to become familiar with a package like gt, allowing for nearly limitless customization. A few methods for creating and modifying gt tables are shown below, but more are found in [package documentation](https://gt.rstudio.com/articles/gt.html).

### Corpus details

By default, a corpus prepared by tmtyro will `tabulize()` into a table showing word counts for each document. A simple version of this can be prepared by hand with very little effort:

```{r}
gt_details <- corpus_dubliners |> 
  count(doc_id) |> 
  gt()

gt_details
```

Once the table is prepared, gt allows for further tweaking---for instance, to format word counts for readability, hide the `doc_id` column header, and rename `n` as `words`:

```{r}
gt_details |> 
  fmt_integer(n) |> 
  cols_label(
    doc_id = "",
    n = "words")
```


### Word frequencies

The standard workflow for preparing a polished table of high-frequency word counts with tmtyro---`add_frequency() |> tabulize()`---will easily show a few of the most used words in each document. To use `get_frequency()` when adding columns for word counts, a chain of functions will prepare a summary table---`group_by() |> summarize() |> ungroup() |> slice_max()`. Once it's ready, `gt()` will do the rest.

```{r}
#| message: false
dubliners_count <- dubliners_count |>
  group_by(doc_id, word) |> 
  summarize(n = max(n)) |> 
  ungroup() |> 
  slice_max(
    order_by = n, 
    by = doc_id, 
    n = 3) # show three words each

gt_counts <- dubliners_count |> 
  # limit to three stories for a shorter display
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  gt()

gt_counts
```

The `cols_label()` function from gt can adjust headers, and tmtyro's `collapse_rows()` function cleans up repeated values in the first column: 

```{r}
gt_counts |> 
  cols_label(doc_id = "") |> 
  collapse_rows(doc_id)
```

Choosing to adjust things manually introduces a steeper learning curve, but it also allows for greater customization:

```{r}
dubliners_count |> 
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  gt(groupname_col = "doc_id") |> 
  cols_label(
    word = "") |> 
  data_color(columns = n, palette = "PuBuGn") |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups())
```

Dictionary matches, including for sentiment, follow the same pattern.

### Vocabulary richness

A similar manual workflow can be used to prepare tables of vocabulary richness:

```{r}
gt_vocab <- corpus_dubliners |> 
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  group_by(doc_id) |> 
  summarize(
    length = n(),
    vocab_count = sum(is_new(word)),
    ttr = last(get_ttr(word)),
    hapax_count = sum(is_hapax(word)),
    htr = last(get_hir(word))) |> 
  ungroup() |> 
  gt()

gt_vocab
```

Here, tab spanners can be added to approximate the version created by a typical tmtyro workflow:

```{r}
gt_vocab |> 
  tab_spanner(
    label = "vocabulary",
    columns = c("vocab_count", "ttr")) |> 
  tab_spanner(
    label = "hapax",
    columns = c("hapax_count", "htr")) |> 
  cols_label(
    vocab_count = "total",
    ttr = "ratio",
    hapax_count = "total",
    htr = "ratio") |> 
  fmt_number(c(ttr, htr), decimals = 3)
```

### Extending `tabulize()`

Learning some of gt's functions can also be helpful for customizing outputs derived from `tabulize()`, which are just gt objects. To demonstrate the usefulness of customization, we can create a short table of word counts by selecting a few titles and showing the default output from `tabulize()`:

```{r}
some_docs <- unique(corpus_dubliners$doc_id)[c(1:3, 12, 15)]

gt_1 <- corpus_dubliners |> 
  filter(doc_id %in% some_docs) |> 
  tabulize()

gt_1
```

This table is functional but not necessarily pretty. While beauty is subjective, customization makes it possible to aim for something clean like this:

```{r}
gt_1 |> 
  tab_style(
    style = cell_borders(
      sides = "all", 
      color = NULL),
    locations = cells_body()) |> 
  tab_style(
    style = cell_text(size = pct(70)),
    locations = cells_column_labels()
  ) |> 
  cols_align(
    align = "right",
    columns = doc_id) |> 
  opt_css(
    css = ".gt_col_headings {border-bottom-color: #FFFFFF !important;}"
  )
```

## Preparing figures with ggplot2

Vectorized functions beginning `get_...()` and `is...()` won't work well with tmtyro's `visualize()` function, which is made for the standard workflow. Those visualizations are all built with ggplot2, so they can be recreated with a little effort and nearly limitless customization. Consult the [ggplot2 documentation](https://ggplot2.tidyverse.org/articles/ggplot2.html) to learn more methods than can be demonstrated here.

### Corpus details

By default, a corpus prepared with `load_texts()` will `visualize()` into a bar chart showing word counts for each document. Preparing something manually is pretty simple, even if it doesn't compare well to the default output:

```{r fig.width = 5, fig.height = 5, fig.show = "hold", out.width  =  "50%"}
# default output
visualize(corpus_dubliners)

# manual output
corpus_dubliners |> 
  count(doc_id) |> 
  ggplot(aes(
    x = n, 
    y = doc_id)) +
  geom_col()
```

Among other things, `visualize()` preserves the order of documents from top to bottom, adjusts labeling, and adds some settings for theme and color. Alone, each is a simple change. But everything adds up when polishing publication-ready graphs, including customizing gridlines, adjusting label spacing, and formatting numbers:

```{r}
corpus_dubliners |> 
  count(doc_id) |> 
  # reverse doc_id order
  mutate(doc_id = forcats::fct_rev(doc_id)) |> 
  ggplot(aes(
    x = n, 
    y = doc_id, 
    # add color
    fill = doc_id)) +
  geom_col(show.legend = FALSE) +
  # adjust number format and shift y-axis labels
  scale_x_continuous(
    labels = scales::label_comma(),
    expand = c(0, 0)) +
  # change the theme background
  theme_minimal() +
  # adjust labels
  labs(
    x = "length (words)",
    y = NULL) +
  # adjust grid lines
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())
```

### Word frequencies

When used after `add_frequency()`, `visualize()` will prepare a faceted graph of some of the top word frequencies for each document. To create something similar manually, using `mutate()` with a vectorized function like `get_frequency()` or `get_tf_by()`, it's necessary to prepare a table with `summarize()` and `slice_max()` before piping it to `ggplot()`:

```{r}
#| message: false
corpus_dubliners |>
  mutate(
    n = get_tf_by(word, doc_id)) |> 
  group_by(doc_id, word) |> 
  summarize(n = max(n)) |> 
  ungroup() |> 
  slice_max(
    order_by = n, 
    by = doc_id, 
    n = 3) |> # show 3 words each
  ggplot(aes(n, word)) +
  geom_col() +
  facet_wrap(vars(doc_id), scales = "free")
```

The resulting graph can be further customized with ggplot2's functions.

### Vocabulary richness

The default visualizations tmtyro makes after `add_vocabulary()` are highly customized. It isn't hard to make a simple version after a vectorized function like `get_cumulative_vocabulary()`, but this version can lack readability:

```{r}
corpus_dubliners |> 
  group_by(doc_id) |> 
  mutate(
    vocab = get_cumulative_vocabulary(word), 
    progress = row_number()) |> 
  ungroup() |> 
  ggplot(aes(
    x = progress, 
    y = vocab, 
    color = doc_id)) +
  geom_line()
```

Adding direct labels is often worth the effort:

```{r}
dubliners_vocab <- corpus_dubliners |> 
  group_by(doc_id) |> 
  mutate(
    vocab = get_cumulative_vocabulary(word), 
    progress = row_number()) |> 
  ungroup()

# table of labels and locations
document_labels <- dubliners_vocab |> 
  group_by(doc_id) |> 
  summarize(
    vocab = last(vocab),
    progress = last(progress)) |> 
  ungroup()

dubliners_vocab |> 
  ggplot(aes(
    x = progress, 
    y = vocab, 
    color = doc_id)) +
  geom_line() +
  geom_point(
    data = document_labels) +
  # avoid overlapping labels
  ggrepel::geom_text_repel(
    data = document_labels,
    aes(label = doc_id)) +
  theme(legend.position = "none")
```


