---
title: "Customizing tables"
subtitle: "Leveraging gt and tinytable"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tmtyro)
library(gt)
```

The standard tmtyro workflow, with functions like `add_vocabulary()` and `add_sentiment()`, works easily with `tabulize()` to generate clean, useful tables that communicate results effectively. These tables are designed to help users focus on their work without needing to worry about formatting, presentation, or code. For those advancing beyond the tyro stage, learning to customize this output or even to create tables from scratch can be a valuable next step.

## Starting from `tabulize()`

The tables tmtyro creates offer a good starting point for anyone interested in learning more about gt and related packages. Since `tabulize()` creates standard gt tables, they can be modified using standard functions from that package or from extension packages like gtExtras.

### Alignment

By default, character columns in tables prepared by `tabulize()` (and gt in general) are center aligned. To adjust this alignment, use gt's `cols_align()` function.

```{r}
#| message: false
library(dplyr)
library(gt)
library(tmtyro)
corpus_dubliners <- get_gutenberg_corpus(2814) |> 
  load_texts() |> 
  identify_by(part) |> 
  standardize_titles() |> 
  select(doc_id, word)

# Choose just 5 stories
some_docs <- unique(corpus_dubliners$doc_id)[c(1:3, 12, 15)]

corpus_dubliners <- corpus_dubliners |> 
  filter(doc_id %in% some_docs)

# tabulize() typically centers text columns
corpus_dubliners |> 
  tabulize()

# cols_align() adjusts alignment
corpus_dubliners |> 
  tabulize() |> 
  cols_align(
    align = "left",
    columns = doc_id)
```


### Themes

Outputs can be highly customized using themes built into packages like gtExtras. 

```{r}
library(gtExtras)
corpus_dubliners |> 
  tabulize() |> 
  gt_theme_excel()
```

Many theme options are available, adjusting coloring, font face, and text size. They're easy to add with functions beginning `gt_theme_...()`:

```{r}
corpus_dubliners |> 
  tabulize() |> 
  gt_theme_538()
```

For more theme options, see the [gtExtras documentation](https://jthomasmock.github.io/gtExtras/reference/index.html#themes) online.

### Titles and Summary Rows

The examples shown here barely scratch the surface of options available with gt. Summary rows, added with `grand_summary_rows()` make it easy to share corpus statistics. Titles and subtitles, added with `tab_header()`, help clarify conclusions and a main takeaways:

```{r}
corpus_dubliners |> 
  tabulize() |>
  cols_align("left", columns = doc_id) |> 
  grand_summary_rows(
    fns = list("avg" ~ mean(.) |> 
                 scales::label_comma(accuracy = 0.1)()), 
    columns = "n") |> 
  tab_header(
    title = md("Word counts in *Dubliners* stories"),
    subtitle = "“The Dead” is about three times the average length.") |> 
  opt_align_table_header("left")
```

### Going further

Combining these methods with those explained in greater depth in gt's documentation can allow for truly customized tables. The default table, for instance, is functional but not necessarily pretty. Customization makes it possible to aim for something clean like this:

```{r}
corpus_dubliners |> 
  tabulize() |> 
  tab_style(
    style = cell_borders(
      sides = "all", 
      color = NULL),
    locations = cells_body()) |> 
  tab_style(
    style = cell_text(size = pct(70)),
    locations = cells_column_labels()
  ) |> 
  cols_align(
    align = "right",
    columns = doc_id) |> 
  opt_css(
    css = ".gt_col_headings {border-bottom-color: #FFFFFF !important;}"
  )
```

## Starting with gt

tmtyro's `tabulize()` only works with a standard workflow using functions like `add_vocabulary()` and `add_frequency()`. Preparing similar tables manually is possible with familiarity with packages like gt or tinytable. A few methods for creating and modifying gt tables are shown below, but more are found in [package documentation](https://gt.rstudio.com/articles/gt.html). 

### Corpus details

By default, a corpus prepared by tmtyro will `tabulize()` into a table showing word counts for each document. A simple version of this can be prepared by hand with very little effort:

```{r}
gt_details <- corpus_dubliners |> 
  count(doc_id) |> 
  gt()

gt_details
```

Once the table is prepared, gt allows for further tweaking---for instance, to format word counts for readability, hide the `doc_id` column header, and rename `n` as `words`:

```{r}
gt_details |> 
  fmt_integer(n) |> 
  cols_label(
    doc_id = "",
    n = "words")
```

### Word frequencies

The standard workflow for preparing a polished table of high-frequency word counts with tmtyro---`add_frequency() |> tabulize()`---will easily show a few of the most used words in each document. To use `get_frequency()` when adding columns for word counts, a chain of functions will prepare a summary table---`distinct() |> slice_max()`. Once it's ready, `gt()` will do the rest.

```{r}
#| message: false
dubliners_count <- corpus_dubliners |>
  group_by(doc_id) |> 
  mutate(
    n = get_frequency(word)) |> 
  ungroup() |> 
  distinct() |> 
  slice_max(
    order_by = n, 
    by = doc_id, 
    n = 3) # show three words each

gt_counts <- dubliners_count |> 
  # limit to three stories for a shorter display
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  gt()

gt_counts
```

The `cols_label()` function from gt can adjust headers, and tmtyro's `collapse_rows()` function hides repeated values in a column: 

```{r}
gt_counts |> 
  cols_label(doc_id = "") |> 
  collapse_rows(doc_id)
```

Choosing to adjust things manually introduces a steeper learning curve, but it also allows for greater customization:

```{r}
dubliners_count |> 
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  gt(groupname_col = "doc_id") |> 
  cols_label(
    word = "") |> 
  data_color(columns = n, palette = "PuBuGn") |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups())
```

Dictionary matches, including for sentiment, follow the same pattern.

### Vocabulary richness

A similar manual workflow can be used to prepare tables of vocabulary richness. Without customization, `gt()` prepares a table that isn't as clear as it could be:

```{r}
dubliners_vocab <- corpus_dubliners |> 
  filter(doc_id %in% c("The Sisters", "An Encounter", "The Dead")) |> 
  group_by(doc_id) |> 
  summarize(
    words = n(),
    vocab_count = sum(is_new(word)),
    ttr = last(get_ttr(word)),
    hapax_count = sum(is_hapax(word)),
    htr = last(get_hir(word))) |> 
  ungroup()

gt_vocab <- dubliners_vocab |> 
  gt()

gt_vocab
```

Here, tab spanners can be added to approximate the version created by a typical tmtyro workflow:

```{r}
gt_vocab |> 
  tab_spanner(
    label = "vocabulary",
    columns = c("vocab_count", "ttr")) |> 
  tab_spanner(
    label = "hapax",
    columns = c("hapax_count", "htr")) |> 
  cols_label(
    vocab_count = "total",
    ttr = "ratio",
    hapax_count = "total",
    htr = "ratio") |> 
  fmt_number(c(ttr, htr), decimals = 3)
```

## Starting with tinytable

Of course, many other options exist in R for preparing tables to communicate findings. One of these, `tt()` from the tinytable package, is worth consideration. A few methods for preparing tinytable tables are shown here, but more are found in [package documentation](https://vincentarelbundock.github.io/tinytable/). 

### Corpus details

The standard function for using tinytable is `tt()`:

```{r}
library(tinytable)
details_tt <- corpus_dubliners |> 
  count(doc_id) |> 
  tt()

details_tt
```

Adjusting this output is straightforward using a few functions that use a standard syntax. Each references rows with the argument "i" and columns with the argument "j". Data format is adjusted using `format_tt()`, and output style is modified with `style_tt()`. For instance, to change the number format in the "n" column shown here, use `format_tt()` like this:

```{r}
details_tt <- details_tt |> 
  format_tt(
    j = 2,
    digits = 0,
    num_mark_big = ",")

details_tt
```

Column names are adjusted using the standard `colnames()` or `setNames()` functions from R:

```{r}
colnames(details_tt) <- c("", "words")

details_tt
```

Properties like column alignment can be adjusted with `style_tt()`:

```{r}
details_tt |> 
  style_tt(
    j = 2,
    align = "r"
  )
```

### Word frequencies

While tmtyro offers `collapse_rows()` to limit repeated values in gt tables, these need to be suppressed manually using tinytable's `rowspan` argument in `style_tt()`:

```{r}
dubliners_count |> 
  group_by(doc_id) |> 
  slice_head(n = 3) |> 
  ungroup() |> 
  tt() |> 
  style_tt(
    i = c(1, 4, 7, 10, 13), 
    j = 1, 
    rowspan = 3, 
    alignv = "t")
```

Unfortunately, this process of manually indicating rows is fiddly and prone to error. Any miscount will make the table misrepresent the data. As an alternative, consider adjusting the underlying table before using `tt()` to cut out repeating values, using `mutate()`, `case_when()`, and `lag()`:

```{r}
dubliners_count |> 
  group_by(doc_id) |> 
  slice_head(n = 3) |> 
  ungroup() |> 
  mutate(
    doc_id = case_when(
      doc_id == lag(doc_id) ~ "",
      TRUE ~ doc_id
    )) |> 
  tt()
```

Alternatively, use automatic grouping, indicating rows with `group_tt()`:

```{r}
count_table <- dubliners_count |> 
  group_by(doc_id) |> 
  slice_head(n = 3) |> 
  ungroup()

# Drop the doc_id column with select(), then reference it in group_tt()
my_tt <- count_table |> 
  select(-doc_id) |> 
  tt() |> 
  group_tt(i = as.character(count_table$doc_id))

my_tt
```

To format these group rows, we can use the attribute `my_tt@group_index_i` to get the row numbers:

```{r}
my_tt |> 
  style_tt(
    i = my_tt@group_index_i, 
    bold = TRUE,
    background = "lightgreen")
```

### Vocabulary richness

Tables for reporting vocabulary richness often need a lot of customizing. By default, `tt()` prepares a table that leaves a lot to be desired: 

```{r}
tt_vocab <- dubliners_vocab |> 
  tt()

tt_vocab
```

Among other things, we might want to adjust number formatting with `format_tt()`, set alignment with `style_tt()`, rename columns using `colnames()` or `setNames()`, and add labels over column groupings with `group_tt()`:

```{r}
tt_vocab |> 
  group_tt(
    j = list(
      "vocabulary" = 3:4,
      "hapax" = 5:6)) |> 
  format_tt(
    j = c(2:3, 5),
    digits = 0,
    num_mark_big = ",") |> 
  style_tt(
    j = c(2:3, 5),
    align = "r") |> 
  format_tt(
    j = c(4, 6),
    digits = 3,
    num_fmt = "decimal",
    num_zero = TRUE) |> 
  setNames(c("", "words", "total", "ratio", "total", "ratio"))
```

In the end, none of this is overwhelming, and results can be clearly prepared for communication.
